---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 4
)
```

# CSHShydRometry

<!-- badges: start -->
[![R-CMD-check](https://github.com/CSHS-CWRA/CSHShydRometry/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CSHS-CWRA/CSHShydRometry/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/CSHS-CWRA/CSHShydRometry/graph/badge.svg)](https://app.codecov.io/gh/CSHS-CWRA/CSHShydRometry)
[![License:
MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://cran.r-project.org/web/licenses/MIT)
<!-- badges: end -->

A rating curve turns a river's stage (its water level, which is easier to
record continuously) into discharge (the flow, which is not). It is fitted to
gaugings: occasions when both were measured. This package fits rating curves
by a range of statistical methods, and calculates uncertainty associated with
them.

## Installation

The package is not on CRAN yet. Get it from GitHub:

``` r
remotes::install_github("CSHS-CWRA/CSHShydRometry")
```

## Fitting a single-segment rating curve

```{r}
library(CSHShydRometry)
```

The package comes with 93 gaugings from the Thompson River (Water Survey of
Canada station 08LF051). `stage` is in meters and `discharge` in cubic meters
per second. A few gaugings also carry a reported uncertainty.

```{r}
head(thompson)
```

By convention, rating curves are drawn with stage on the vertical axis.
Discharge rises faster than stage, so the points bend over to the right:

```{r plot-data}
plot(stage ~ discharge, data = thompson)
```

The classic rating curve is a power law relating discharge $Q$ to stage $h$:

$$
Q = a (h - c)^b
$$

`rc_nls()` is one method, which fits this relationship by nonlinear least squares:

```{r fit}
fit <- rc_nls(discharge, stage, data = thompson)
```

Like the result of `lm()` or `glm()`, `fit` is a fitted-model object. Printing
it says what kind of model it is, and it works with the usual tools, such as
`coef()` and `predict()`. `?rating_curve` describes what is inside.

```{r fit-print}
fit
coef(fit)
```

Here $c$ is the stage at which the flow would stop, $b$ says how quickly the
flow grows as the water rises above that, and $a$ sets the scale: it is the
discharge when the water is one meter above $c$.

## Predicting discharge

`predict()` evaluates the curve at the stages you give it:

```{r predict}
predict(fit, stage = c(1, 3, 6))
```

Ask for a confidence level to get limits for the curve itself, and a
prediction level to get limits for a new gauging:

```{r limits}
predict(fit, stage = c(1, 3, 6), conflev = 0.95, predlev = 0.95)
```

Leave out `stage` to cover the whole range of the gaugings, which is handy for
plotting:

```{r band}
band <- predict(fit, conflev = 0.95, predlev = 0.95)

plot(stage ~ discharge, data = thompson)
lines(stage ~ fit, data = band)
lines(stage ~ pi_lwr, data = band, lty = 2)
lines(stage ~ pi_upr, data = band, lty = 2)
```

The dashed lines are the 95% prediction limits. The confidence limits are
there too, in `ci_lwr` and `ci_upr`, but on this scale they sit almost on the
curve.

## Other models

Every model has an `rc_*()` function and a `predict()` method:

- `rc_nls()`: power law, fitted on the natural scale.
- `rc_log_ols()`, `rc_log_nls()`: power law, fitted on the log–log scale.
- `rc_gnls()`: power law, with the scatter estimated as a power of the flow.
- `rc_poly()`, `rc_loess()`: a polynomial, or a smooth curve.
- `rc_2seg_nls()`: two power laws joined at a breakpoint (more below).

Swapping one model for another changes one line:

```{r poly}
fit_poly <- rc_poly(discharge, stage, data = thompson)
predict(fit_poly, stage = c(1, 3, 6))
```

`predict()` always gives back the same columns, so results from different
models can be stacked with `rbind()`:

```{r stack}
rbind(
  predict(fit, stage = 3, conflev = 0.95),
  predict(fit_poly, stage = 3, conflev = 0.95)
)
```

## Two-segment curves

Where the river's control changes (say, when the water rises out of the
channel and over a floodplain), one power law is not enough. The Ardèche at
Sauze, from the RBaM package, is such a river. RBaM calls its columns `H`,
`Q` and `uQ`; here we give them the names used in this package. Each gauging
comes with its standard uncertainty, in cubic meters per second:

```{r sauze}
sauze <- data.frame(
  stage = RBaM::SauzeGaugings$H,
  discharge = RBaM::SauzeGaugings$Q,
  uncertainty_sd = RBaM::SauzeGaugings$uQ
)
head(sauze)
```

Below about 1 m the stage climbs steeply with discharge; above about 2 m it
climbs much more slowly. There are no gaugings in between, so where the control
changes has to be estimated:

```{r plot-sauze}
plot(stage ~ discharge, data = sauze)
```

`rc_2seg_nls()` fits two power laws that meet at a breakpoint, $k$. Here we
weight each gauging by its reported uncertainty:

```{r twoseg}
fit2 <- rc_2seg_nls(discharge, stage, data = sauze,
                    wts_code = "spec", wts = 1 / uncertainty_sd^2)
fit2$pars$k
```

The fit is sensitive to where the breakpoint search starts, so by default it
tries several starting points and keeps the best.

### Limits near the breakpoint

For two-segment curves, `predict()` offers two ways to compute the limits.
The default, `method = "delta"`, is fast but unreliable near the breakpoint:
the curve has a corner there, and the method's straight-line approximation
jumps across it. `method = "boot"` refits the curve to resampled gaugings;
it is slow, but it behaves at the breakpoint.

Just either side of the breakpoint, the delta band jumps and the bootstrap
band does not:

```{r delta-boot}
near_k <- fit2$pars$k + c(-0.05, 0.05)
delta <- predict(fit2, stage = near_k, conflev = 0.95)
boot <- predict(fit2, stage = near_k, conflev = 0.95, method = "boot", B = 200,
                seed = 1)

delta$ci_upr - delta$ci_lwr
boot$ci_upr - boot$ci_lwr
```

In a simulation study, a nominal 95% delta interval covered the true curve
only about two-thirds of the time just above the breakpoint, against roughly
97% for the bootstrap. Away from the breakpoint the two agree. So use
`"delta"` for a quick look, and `"boot"` where the interval matters.

## Acknowledgements

The name of this R package is in recognition of the support provided by the
[Canadian Society for Hydrological Sciences (CSHS)](https://cwra.org/en/affiliates-programs/cshs/)
which is an affiliated society of the Canadian Water Resources Association (CWRA).

## Contributing

Contributions are welcome; see [CONTRIBUTING.md](CONTRIBUTING.md). Please note
that this project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this
project you agree to abide by its terms.
