---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 4
)
```

# CSHShydRometry

<!-- badges: start -->
[![R-CMD-check](https://github.com/CSHS-CWRA/CSHShydRometry/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CSHS-CWRA/CSHShydRometry/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/CSHS-CWRA/CSHShydRometry/graph/badge.svg)](https://app.codecov.io/gh/CSHS-CWRA/CSHShydRometry)
[![License:
MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://cran.r-project.org/web/licenses/MIT)
<!-- badges: end -->

A rating curve turns a river's stage (its water level, which is easier to
record continuously) into discharge (the flow, which is not). It is fitted to
gaugings: occasions when both were measured. This package fits rating curves
by a range of statistical methods, and calculates uncertainty associated with
them.

## Installation

The package is not on CRAN yet. Get it from GitHub:

``` r
remotes::install_github("CSHS-CWRA/CSHShydRometry")
```

## A first rating curve

The package comes with 93 gaugings from the Thompson River (Water Survey of
Canada station 08LF051), with `stage` in meters and `discharge` in cubic
meters per second. Rating curves are drawn with stage on the vertical axis:

```{r plot-data}
library(CSHShydRometry)
plot(stage ~ discharge, data = thompson)
```

The classic rating curve is a power law, $Q = a (h - c)^b$, relating
discharge $Q$ to stage $h$, where $c$ is the offset: the stage at which the
flow would stop. `rc_powerlaw()` fits it by least squares. These gaugings
scatter more at higher flows, so we let the standard deviation of the
scatter grow in proportion to the flow, with `variance = "prop"`:

```{r fit}
fit <- rc_powerlaw(discharge, stage, data = thompson, variance = "prop")
coef(fit)
```

## Confidence and prediction limits

`predict()` evaluates the curve at the stages you give it. Ask for a
confidence level to get limits for the curve itself, and a prediction level
to get limits for a new gauging:

```{r limits}
predict(fit, new_stage = c(1, 3, 6), conflev = 0.95, predlev = 0.95)
```

Leave out `new_stage` to cover the range of the gaugings, which is handy for
plotting:

```{r band}
band <- predict(fit, conflev = 0.95, predlev = 0.95)
plot(stage ~ discharge, data = thompson, col = "grey50")
lines(stage ~ fit, data = band)
lines(stage ~ pi_lwr, data = band, lty = 2)
lines(stage ~ pi_upr, data = band, lty = 2)
```

The dashed lines are the 95% prediction limits. The confidence limits, in
`ci_lwr` and `ci_upr`, sit almost on the curve at this scale, and are not shown.

## Other models

Every model has an `rc_*()` function, and swapping one for another changes
one line:

- `rc_powerlaw()`: a power law, fitted on the original scale.
- `rc_powerlaw_log()`: a power law, fitted on the log-log scale.
- `rc_poly()`, `rc_loess()`: a polynomial, or a smooth curve.
- `rc_2seg_powerlaw()`: two power laws joined at a breakpoint (below).

`predict()` gives back the same columns for every model, so results from
different models line up:

```{r}
fit_poly <- rc_poly(
  discharge,
  stage,
  data = thompson,
  variance = "prop"
)
predict(fit, new_stage = 3, conflev = 0.95)
predict(fit_poly, new_stage = 3, conflev = 0.95)
```

## Two-segment curves

Where the river's control changes, say when the water rises out of the
channel onto a floodplain, one power law is not enough. `rc_2seg_powerlaw()`
fits two power laws that meet at a breakpoint stage, estimated along with
them. The Ardèche at Sauze, from the RBaM package, is such a river; each of
its gaugings comes with a reported standard uncertainty, which gives its
variance:

```{r twoseg}
sauze <- tibble::tibble(
  stage = RBaM::SauzeGaugings$H,
  discharge = RBaM::SauzeGaugings$Q,
  uncertainty_sd = RBaM::SauzeGaugings$uQ
)
fit2 <- rc_2seg_powerlaw(
  discharge, stage,
  data = sauze,
  variance = var_spec(sauze$uncertainty_sd^2)
)

plot(stage ~ discharge, data = sauze, col = "grey50")
lines(stage ~ fit, data = predict(fit2))
abline(h = fit2$curve_parameters$k, lty = 3)
```

The dotted line marks the estimated breakpoint. The curve has a corner
there, so for limits near it, use
`predict(fit2, conflev = 0.95, method = "boot")`: the default delta method is
fast, but unreliable at the corner.

## Learn more

- `vignette("fitting")`: choosing gaugings, looking at their scatter, and
  every way the package fits a curve.
- `vignette("uncertainty")`: confidence and prediction limits, what they
  assume, and where they mislead.

## Related software

Most agencies build ratings in commercial hydrometric data systems:
[AQUARIUS Time-Series](https://aquaticinformatics.com/) (Aquatic
Informatics), used by the Water Survey of Canada and the USGS, and
[WISKI](https://www.kisters.net/) and
[Hydstra](https://www.kisters.net/hydstra/) (KISTERS). They manage the whole
workflow, from gaugings to published discharge, but are proprietary.

Open-source tools for fitting rating curves are mostly Bayesian:

| Software | Language | Approach |
|---|---|---|
| [BaRatin / BaRatinAGE](https://github.com/BaRatin-tools/BaRatinAGE) (INRAE) | Fortran, with a graphical interface | Bayesian; priors from the hydraulic controls; multi-segment curves through a configuration matrix |
| [RBaM](https://CRAN.R-project.org/package=RBaM) (INRAE) | R | R interface to BaM, the engine behind BaRatin |
| [bdrc](https://CRAN.R-project.org/package=bdrc) | R | Bayesian hierarchical power law and generalized power law, with constant or varying scatter |
| [ratingcurve](https://github.com/thodson-usgs/ratingcurve) (USGS, provisional) | Python | Bayesian segmented power law, with PyMC |

CSHShydRometry instead fits by least squares, with no priors to specify. Its
limits come from the delta method or a bootstrap, and it has explicit
schemes for how the scatter changes with the flow. That makes it quick to
use, and a classical reference against which to compare the Bayesian
methods. Beware that notation differs between tools: BaRatin, for example,
writes the power law as $Q = a(h - b)^c$, swapping the roles of $b$ and $c$
used here. For a comparison of methods for rating-curve uncertainty, see
Kiang et al. (2018), "A comparison of methods for streamflow uncertainty
estimation", *Water Resources Research*,
[doi:10.1029/2018WR022708](https://doi.org/10.1029/2018WR022708).

## Citation

To cite CSHShydRometry in publications, please use the following (which
`citation("CSHShydRometry")` also prints):

```{r citation, echo = FALSE, results = "asis"}
cit <- citation("CSHShydRometry")
cat(format(cit, style = "text"), "\n\nAs BibTeX:\n\n")
cat("``` bibtex", format(cit, style = "bibtex"), "```", sep = "\n")
```

## Acknowledgements

The name of this R package is in recognition of the support provided by the
[Canadian Society for Hydrological Sciences (CSHS)](https://cwra.org/en/affiliates-programs/cshs/),
which is an affiliated society of the Canadian Water Resources Association (CWRA).

## Contributing

Contributions are welcome; see [CONTRIBUTING.md](CONTRIBUTING.md). Please note
that this project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this
project you agree to abide by its terms.
