---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CSHShydRometry

<!-- badges: start -->
[![R-CMD-check](https://github.com/CSHS-CWRA/CSHShydRometry/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/CSHS-CWRA/CSHShydRometry/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/CSHS-CWRA/CSHShydRometry/graph/badge.svg)](https://app.codecov.io/gh/CSHS-CWRA/CSHShydRometry)
<!-- badges: end -->

Frequentist methods for fitting stage–discharge rating curves, with confidence
and prediction limits.

Derived from work by Dan Moore on rating-curve methods, restructured into a
model–predict form: every model is fitted by a `rc_*()` constructor and
summarised by a `predict()` method.

## Installation

``` r
remotes::install_github("CSHS-CWRA/CSHShydRometry")
```

## Fitting a curve

```{r fit}
library(CSHShydRometry)

fit <- rc_nls(q, h, data = thompson)
fit

predict(fit, hpred = c(1, 3, 6), conflev = 0.95, predlev = 0.95)
```

Leave out `hpred` and `predict()` evaluates the curve over the observed stage
range, which is handy for plotting:

```{r plot, fig.height = 4}
band <- predict(fit, conflev = 0.95, predlev = 0.95)
plot(q ~ h, data = thompson, pch = 16, col = "grey40",
     xlab = "Stage (m)", ylab = expression(Discharge ~ (m^3 ~ s^-1)))
polygon(c(band$h, rev(band$h)), c(band$pi_lwr, rev(band$pi_upr)),
        col = adjustcolor("steelblue", 0.2), border = NA)
polygon(c(band$h, rev(band$h)), c(band$ci_lwr, rev(band$ci_upr)),
        col = adjustcolor("steelblue", 0.4), border = NA)
lines(fit ~ h, data = band, col = "steelblue", lwd = 2)
```

Single-segment models: `rc_log_ols()` and `rc_log_nls()` fit a power law on the
log–log scale, `rc_nls()` on the natural scale, `rc_gnls()` estimates the error
variance as a power of the mean, and `rc_poly()` and `rc_loess()` offer
non-power-law alternatives.

Two-segment curves are fitted by `rc_nls_2seg()`, joined either continuously
(`config = "piecewise"`) or additively (`config = "compound"`) at an estimated
breakpoint.

## Consistent output

`predict()` returns the same columns whatever the model, method or weighting:
`h`, `fit`, and — when asked for — `ci_lwr`/`ci_upr` and `pi_lwr`/`pi_upr`.
Quantities that cannot be computed come back as `NA` rather than as missing
columns, so a batch of approaches stacks directly:

```{r stack}
fits <- list(
  log_ols = rc_log_ols(q, h, data = thompson),
  nls     = rc_nls(q, h, data = thompson),
  nls_cv  = rc_nls(q, h, data = thompson, wts_code = "prop"),
  poly    = rc_poly(q, h, data = thompson),
  loess   = rc_loess(q, h, data = thompson)
)
do.call(rbind, lapply(names(fits), function(nm) {
  cbind(model = nm, predict(fits[[nm]], hpred = 3, conflev = 0.95))
}))
```

## Weighting

Constructors taking `wts_code` offer three error models:

| `wts_code` | assumption |
|---|---|
| `"none"` | constant variance |
| `"prop"` | constant coefficient of variation, fitted by IRLS |
| `"spec"` | variances supplied by the user, e.g. from reported gauging uncertainties |

Under `"spec"` a new observation's scatter is not identified by the fit, so
prediction limits are returned as `NA`.

## Intervals for two-segment curves

`predict.rc_nls_2seg()` takes a `method`:

| method | notes |
|---|---|
| `"delta"` (default) | linearised, fast, **unreliable near the breakpoint** |
| `"boot"` | resamples the gaugings and refits; slow, but trustworthy at the breakpoint |

The default warrants a word. A two-segment mean function is not differentiable
at the breakpoint, so the delta method's linearisation switches form there and
the interval jumps. On one fitted curve the band widened from 29 to
242 m³ s⁻¹ across the breakpoint, and in a simulation study a nominal 95%
delta interval covered the true curve only about two-thirds of the time just
above it, against roughly 97% for the bootstrap. Away from the breakpoint the
delta method behaves normally.

You can see it at Sauze, either side of the fitted breakpoint, with weights
from the reported gauging uncertainties:

```{r twoseg}
sauze <- RBaM::SauzeGaugings
fit2 <- rc_nls_2seg(Q, H, data = sauze, kstart = 1,
                    wts_code = "spec", wts = 1 / sauze$uQ^2)
k <- fit2$pars[["k"]]
k

hp <- k + c(-0.05, 0.05)
delta <- predict(fit2, hpred = hp, conflev = 0.95)
boot <- predict(fit2, hpred = hp, conflev = 0.95, method = "boot",
                B = 200, seed = 1)
data.frame(
  h = hp,
  delta_width = delta$ci_upr - delta$ci_lwr,
  boot_width = boot$ci_upr - boot$ci_lwr
)
```

So `"delta"` for a quick look, and `"boot"` where the interval matters.

## Data

`thompson` ships with the package: 93 gaugings from Water Survey of Canada
station 08LF051, Thompson River. It is close to a single control, so it suits
the single-segment models; its two-segment fit converges only with care.

For the two-segment models the examples and tests prefer the Ardeche at Sauze,
`RBaM::SauzeGaugings`, which has a clear change of control and a reported
uncertainty for every gauging. RBaM is a suggested dependency, used only as a
source of that data — none of the fitting here is Bayesian, and the BaM engine
is not needed.

## Scope

This package covers the frequentist methods only. Bayesian rating-curve
estimation, and a breakpoint-averaged interval method that removes the jump
described above, are deliberately left out for now.

An interval method that draws parameters from their asymptotic normal
distribution and pushes each draw through the model was tried and dropped:
where a segment is poorly identified, the sampled parameters too often give
impossible curves (negative or astronomically large discharges).
