---
title: "Fitting rating curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting rating curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.2,
  out.width = "100%"
)
has_rbam <- requireNamespace("RBaM", quietly = TRUE)
```

This vignette walks through the ways CSHShydRometry fits a rating curve: the
relationship between a river's stage (its water level) and its discharge
(the flow). It starts by looking at the data, which tells you which model to
use, then fits single-segment curves in several ways, and ends with a curve
in two segments. How sure we can be about the fitted curves is the subject
of `vignette("uncertainty")`.

```{r setup}
library(CSHShydRometry)
```

## The gaugings

A rating curve is fitted to gaugings: occasions when both stage and discharge
were measured. The package comes with `r nrow(thompson)` gaugings from the
Thompson River near Spences Bridge, British Columbia, made between
`r format(min(thompson$date), "%Y")` and `r format(max(thompson$date), "%Y")`.

```{r}
thompson
```

Rating curves are drawn with stage on the vertical axis:

```{r}
plot(stage ~ discharge, data = thompson,
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
```

## Which gaugings to use

A rating curve describes the river while its channel stays the same. Channels
change, through erosion, deposition or work on the control, so a curve is
usually fitted to gaugings from one period of stable conditions.

One way to check is to fit a single curve to every gauging and see whether
the gaugings drift away from it over time. The fit here, `rc_power_log()`, is
explained below; its residuals are on the log scale, so 0.05 means about 5%.

```{r}
fit_all <- rc_power_log(discharge, stage, data = thompson)
departure <- residuals(fit_all$model)

plot(thompson$date, 100 * departure,
     xlab = "Date of gauging", ylab = "Departure from the curve (%)")
abline(h = 0, col = "grey")
lines(lowess(thompson$date, 100 * departure, f = 0.3), lwd = 2)
```

The gaugings drift by several percent, up in the 1990s, down in the 2000s,
and so on: as much as their scatter about the curve. The Water Survey's own
rating changed too, three times between 2020 and 2024, as the `rating_table`
column records:

```{r}
table(thompson$rating_table, useNA = "ifany")
```

For a rating to use in practice, you would choose a stable period and keep
only its gaugings, for example with `subset(thompson, rating_table == "11")`.
This vignette uses all `r nrow(thompson)` gaugings to illustrate the methods,
which is fine for that purpose, but the fitted curves average over several
ratings.

## Looking at the scatter

Before choosing a model, look at how the gaugings scatter about a curve.
A loess smooth makes few assumptions about the curve's shape, so its
residuals show the scatter fairly:

```{r, fig.height = 3.4}
smooth <- rc_loess(discharge, stage, data = thompson)
fitted_q <- fitted(smooth$model)
residual_q <- thompson$discharge - fitted_q

op <- par(mfrow = c(1, 2))
plot(fitted_q, residual_q, xlab = "Fitted discharge (m³/s)",
     ylab = "Residual (m³/s)", main = "Residuals")
abline(h = 0, col = "grey")
plot(fitted_q, 100 * residual_q / fitted_q, xlab = "Fitted discharge (m³/s)",
     ylab = "Residual (%)", main = "Relative residuals")
abline(h = 0, col = "grey")
par(op)
```

The residuals fan out as the flow grows, while the relative residuals stay
roughly even. The scatter is roughly proportional to the flow: a constant
percentage, not a constant amount. That matters for the fit, because
ordinary least squares assumes the same scatter at every flow. There are two
ways to allow for it, both below: weight the fit, or fit on the log scale.

## A power law on the original scale

The usual rating curve is a power law, $Q = a (h - c)^b$, where $Q$ is
discharge, $h$ is stage, and $c$ is the stage at which the flow would stop.
`rc_power()` fits it by least squares on the original scale:

```{r}
fit <- rc_power(discharge, stage, data = thompson)
coef(fit)
```

By default every gauging counts equally (`wts_none()`). With
`wts_prop()` the scatter is taken to be proportional to the flow, as the
residuals suggested. The weights then depend on the fitted curve, so the fit
is repeated, reweighting each time, until it settles:

```{r}
fit_prop <- rc_power(discharge, stage, data = thompson, wts = wts_prop())
coef(fit_prop)
fit_prop$irls$iterations
```

`wts_power()` goes one step further and estimates the power of the flow to
which the scatter is proportional:

```{r}
fit_power <- rc_power(discharge, stage, data = thompson, wts = wts_power())
fit_power$wts$exponent
```

An estimated power close to 1 supports `wts_prop()`. The estimate can be
unstable with few gaugings, so compare it with the fit under `wts_prop()`.

When each gauging comes with its own reported uncertainty, `wts_spec()` uses
it directly; there is an example with the two-segment curve below. See
`?wts` for all the weighting schemes.

## A power law on the log scale

Taking logs turns the power law into a straight line in $\log(h - c)$:

$$
\log Q = \log a + b \log(h - c)
$$

Equal scatter on the log scale is scatter proportional to the flow, so
`rc_power_log()` allows for the fanning out by its choice of scale, without
weights:

```{r}
fit_log <- rc_power_log(discharge, stage, data = thompson)
coef(fit_log)
```

Back on the original scale, this curve estimates the geometric mean
discharge at each stage, which is a little below the arithmetic mean.
`fit_log$a_corrected` holds two versions of $a$ corrected for that.

If the stage of zero flow is known, say from a survey of the control, give it
as `zero_flow_stage` and only $a$ and $b$ are estimated:

```{r}
coef(rc_power_log(discharge, stage, data = thompson, zero_flow_stage = -1.3))
```

## Other curve shapes

Not every rating follows a power law. `rc_poly()` fits a polynomial in stage,
and `rc_loess()` a smooth curve with no set shape. Both accept the same
weighting schemes as `rc_power()`.

```{r}
fit_cubic <- rc_poly(discharge, stage, data = thompson, degree = 3)
fit_smooth <- rc_loess(discharge, stage, data = thompson)
```

## Comparing the fits

Every fit has a `predict()` method that returns the same columns, so the
curves can be compared side by side at chosen stages:

```{r}
at <- c(1, 4, 8)
data.frame(
  stage = at,
  power = predict(fit, stage = at)$fit,
  power_prop = predict(fit_prop, stage = at)$fit,
  power_log = predict(fit_log, stage = at)$fit,
  cubic = predict(fit_cubic, stage = at)$fit,
  loess = predict(fit_smooth, stage = at)$fit
)
```

Without `stage`, `predict()` covers the range of the gaugings, which makes it
easy to draw the curves:

```{r}
plot(stage ~ discharge, data = thompson, col = "grey50",
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
lines(stage ~ fit, data = predict(fit_prop), lwd = 2)
lines(stage ~ fit, data = predict(fit_log), lwd = 2, lty = 2)
lines(stage ~ fit, data = predict(fit_cubic), lwd = 2, lty = 3)
legend("bottomright", c("Power law, proportional weights", "Power law, log scale",
       "Cubic polynomial"), lty = 1:3, lwd = 2, bty = "n")
```

Within the range of the gaugings the curves barely differ. They part company
beyond it; `vignette("uncertainty")` takes that up.

## A curve in two segments

Where the control changes, for example when the water rises out of the
channel onto a floodplain, one power law is not enough. `rc_2seg_power()`
fits two power laws that meet at a breakpoint stage, $k$.

The Ardèche at Sauze, from the RBaM package, has such a change. RBaM calls
its columns `H`, `Q` and `uQ`; here they get the names used in this package.
Each gauging comes with its standard uncertainty, in cubic metres per second.

```{r, eval = has_rbam}
sauze <- data.frame(
  stage = RBaM::SauzeGaugings$H,
  discharge = RBaM::SauzeGaugings$Q,
  uncertainty_sd = RBaM::SauzeGaugings$uQ
)
plot(stage ~ discharge, data = sauze,
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
```

The reported uncertainties give the weights directly, through `wts_spec()`.
How the two segments combine above the breakpoint is set by `controls`:
`"successive"` (the default) lets the upper power law take over from the
lower one, and `"additive"` adds it to the discharge the lower one carries.

```{r, eval = has_rbam}
fit_succ <- rc_2seg_power(discharge, stage, data = sauze,
                          wts = wts_spec(1 / uncertainty_sd^2))
fit_add <- rc_2seg_power(discharge, stage, data = sauze,
                         controls = "additive",
                         wts = wts_spec(1 / uncertainty_sd^2))
c(successive = fit_succ$pars$k, additive = fit_add$pars$k)
```

The fit is sensitive to where the search for the breakpoint starts, so by
default it tries several starting values and keeps the best. What each
start led to is recorded:

```{r, eval = has_rbam}
fit_succ$kstart_search
```

```{r, eval = has_rbam}
plot(stage ~ discharge, data = sauze, col = "grey50",
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
lines(stage ~ fit, data = predict(fit_succ), lwd = 2)
lines(stage ~ fit, data = predict(fit_add), lwd = 2, lty = 2)
abline(h = fit_succ$pars$k, col = "grey", lty = 3)
legend("bottomright", c("Successive controls", "Additive controls"),
       lty = 1:2, lwd = 2, bty = "n")
```

Gaugings are sparse near the breakpoint at Sauze, so where exactly it lies is
uncertain, which affects the limits on the curve there; see
`vignette("uncertainty")`.
