---
title: "Fitting rating curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting rating curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.2,
  out.width = "100%"
)
has_rbam <- requireNamespace("RBaM", quietly = TRUE)
```

This vignette walks through the ways CSHShydRometry fits a rating curve: the
relationship between a river's stage (its water level) and its discharge
(the flow). It starts by looking at the data, which tells you which model to
use, then fits single-segment curves in several ways, and ends with a curve
in two segments. How sure we can be about the fitted curves is the subject
of `vignette("uncertainty")`.

```{r setup}
library(CSHShydRometry)
```

## The gaugings

A rating curve is fitted to gaugings: occasions when both stage and discharge
were measured. The package comes with `r nrow(thompson)` gaugings from the
Thompson River near Spences Bridge, British Columbia, made between
`r format(min(thompson$date), "%Y")` and `r format(max(thompson$date), "%Y")`.

```{r}
thompson
```

Rating curves are drawn with stage on the vertical axis:

```{r}
plot(stage ~ discharge, data = thompson,
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
```

## Which gaugings to use

A rating curve describes the river while its channel stays the same. Channels
change, through erosion, deposition or work on the control, so a curve is
usually fitted to gaugings from one period of stable conditions.

One way to check is to fit a model to the entire dataset and see whether
the gaugings systematically miss the curve over time. We'll build that model
here with `rc_powerlaw_log()`, which is explained below. What matters here
are its residuals, which are on the log scale, so multiplied by 100 they are
roughly percentage departures from the curve.

```{r}
fit_all <- rc_powerlaw_log(discharge, stage, data = thompson)
departure <- residuals(fit_all)

plot(thompson$date, 100 * departure,
     xlab = "Date of gauging", ylab = "Departure from the curve (%)")
abline(h = 0, col = "grey")
lines(lowess(thompson$date, 100 * departure, f = 0.3), lwd = 2)
```

The curve's error depends on when the gaugings were made: its predictions
are too low in the 1990s, too high in the 2000s, and so on. The Water
Survey's own rating changed too, as the `rating_table` column records.

For a rating to use in practice, you would choose a stable period and keep
only its gaugings, for example with `subset(thompson, rating_table == "11")`.
This vignette uses all `r nrow(thompson)` gaugings to illustrate the methods,
which is fine for that purpose, but the fitted curves average over several
ratings.

## Looking at the scatter

Before choosing a model, look at how the gaugings scatter about a fitted curve.
A loess smoother makes few assumptions about the curve's shape, so its
residuals show the scatter fairly.

Fit the model and get the fitted values and the residuals.

```{r}
smooth <- rc_loess(discharge, stage, data = thompson)
fitted_q <- fitted(smooth)
residual_q <- residuals(smooth)
```

Now check whether the residuals have an even spread about the curve.

```{r residual-plots, echo = FALSE, fig.height = 3.4}
op <- par(mfrow = c(1, 2))
plot(
  fitted_q,
  residual_q,
  xlab = "Fitted discharge (m³/s)",
  ylab = "Residual (m³/s)",
  main = "Residuals"
)
abline(h = 0, col = "grey")
plot(
  fitted_q,
  100 * residual_q / fitted_q,
  xlab = "Fitted discharge (m³/s)",
  ylab = "Residual (%)",
  main = "Relative residuals"
)
abline(h = 0, col = "grey")
par(op)
```

<details>
<summary>Show the code</summary>

```{r ref.label = "residual-plots", eval = FALSE}
```

</details>

The residuals fan out as the flow grows, while the relative residuals stay
roughly even. The scatter is roughly proportional to the flow: a constant
percentage, not a constant amount. That matters for the fit, because
ordinary least squares doesn't perform as well when the residuals scatter
more with increasing flow.

There are two ways to allow for it, both below: model the variance, or fit
on the log scale.

## A power law on the original scale

The usual rating curve is a power law, $Q = a (h - c)^b$, where $Q$ is
discharge and $h$ is stage. The model parameters are $a$, $b$, and $c$,
with $c$ being the offset: the stage at which the flow would stop.
`rc_powerlaw()` fits this model by least squares on the original scale:

```{r}
fit <- rc_powerlaw(discharge, stage, data = thompson)
coef(fit)
```

To improve the least squares estimation, the `variance` argument lets you
specify how you think the scatter changes with the flow. By default it is
the same at every flow (`var_none()`). With `var_prop()` its standard
deviation is assumed proportional to the flow, which suits the Thompson
gaugings better. Either way, the fit weights each gauging by the reciprocal
of its variance.

```{r}
fit_prop <- rc_powerlaw(
  discharge,
  stage,
  data = thompson,
  variance = var_prop()
)
coef(fit_prop)
```

It's worth knowing about the extra gymnastics this takes. When we say that
the standard deviation of the flow is proportional to the flow itself, we
don't mean the _observed_ flow, but the _average_ flow at a given stage: the
thing being predicted. That's because we're talking about the (vertical)
scatter around the _curve_. So the weights depend on the fitted curve, and
the fitted curve depends on the weights. To deal with this, the fit is
repeated, reweighting each time, until it settles; hence "iteratively
reweighted least squares" (IRLS). Here is the number of iterations it
took:

```{r}
fit_prop$irls$iterations
```

`var_power()` goes one step further than `var_prop()`, and estimates the
power of the flow to which the scatter is proportional:

```{r}
fit_power <- rc_powerlaw(
  discharge,
  stage,
  data = thompson,
  variance = var_power()
)
fit_power$variance$exponent
```

An estimated power close to 1 supports `var_prop()`. The estimate can be
unstable with few gaugings, so compare it with the fit under `var_prop()`.

When each gauging comes with its own reported uncertainty, `var_spec()` uses
it directly; there is an example with the two-segment curve below. See
`?variance` for all the variance schemes.

## A power law on the log scale

Taking logs turns the power law into a straight line in $\log(h - c)$:

$$
\log Q = \log a + b \log(h - c)
$$

Equal scatter on the log scale is scatter proportional to the flow, so
`rc_powerlaw_log()` allows for the fanning out by its choice of scale, with
no variance scheme:

```{r}
fit_log <- rc_powerlaw_log(discharge, stage, data = thompson)
coef(fit_log)
```

Back on the original scale, this curve estimates the geometric mean
discharge at each stage, which is a little below the arithmetic mean.
`fit_log$a_corrected` holds two versions of $a$ corrected for that.

$c$ is the *offset*. For a single power law it is the stage at which the flow
would stop, but with several segments it no longer is, so the package calls
it the offset throughout. If it is known, say from a survey of the control,
give it as `offset` and only $a$ and $b$ are estimated. `rc_powerlaw()` takes
`offset` too.

```{r}
coef(rc_powerlaw_log(discharge, stage, data = thompson, offset = -1.3))
```

## Other curve shapes

Not every rating follows a power law. `rc_poly()` fits a polynomial in stage,
and `rc_loess()` a smooth curve with no set shape. Both accept the same
variance schemes as `rc_powerlaw()`, except `var_power()`.

```{r}
fit_cubic <- rc_poly(discharge, stage, data = thompson, degree = 3)
fit_smooth <- rc_loess(discharge, stage, data = thompson)
```

## Comparing the fits

Every fit has a `predict()` method, so the curves can be compared at chosen
stages:

```{r}
at <- c(1, 4, 8)
predict(fit_prop, new_stage = at)
predict(fit_cubic, new_stage = at)
```

`predict()` returns the same columns for every model, so the results stack.
With the dplyr package (not a dependency of this one), `bind_rows()` can
label each row with its model:

```r
dplyr::bind_rows(
  power_prop = predict(fit_prop, new_stage = at),
  cubic = predict(fit_cubic, new_stage = at),
  .id = "model"
)
#> # A tibble: 6 × 3
#>   model      stage   fit
#>   <chr>      <dbl> <dbl>
#> 1 power_prop     1  252.
#> 2 power_prop     4 1208.
#> 3 power_prop     8 3476.
#> 4 cubic          1  249.
#> 5 cubic          4 1226.
#> 6 cubic          8 3399.
```

Without `new_stage`, `predict()` covers the range of the gaugings, which
makes it easy to draw the curves:

```{r}
plot(stage ~ discharge, data = thompson, col = "grey50",
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
lines(stage ~ fit, data = predict(fit_prop), lwd = 2)
lines(stage ~ fit, data = predict(fit_log), lwd = 2, lty = 2)
lines(stage ~ fit, data = predict(fit_cubic), lwd = 2, lty = 3)
legend("bottomright", c("Power law, var_prop()", "Power law, log scale",
       "Cubic polynomial"), lty = 1:3, lwd = 2, bty = "n")
```

Within the range of the gaugings the curves barely differ. They part company
beyond it; `vignette("uncertainty")` takes that up.

## A curve in two segments

Where the control changes, for example when the water rises out of the
channel onto a floodplain, one power law is not enough. `rc_2seg_powerlaw()`
fits two power laws that meet at a breakpoint stage, $k$.

The Ardèche at Sauze, from the RBaM package, has such a change. RBaM calls
its columns `H`, `Q` and `uQ`; here they get the names used in this package.
Each gauging comes with its standard uncertainty, in cubic meters per second.

```{r, eval = has_rbam}
sauze <- tibble::tibble(
  stage = RBaM::SauzeGaugings$H,
  discharge = RBaM::SauzeGaugings$Q,
  uncertainty_sd = RBaM::SauzeGaugings$uQ
)
plot(
  stage ~ discharge,
  data = sauze,
  xlab = "Discharge (m³/s)",
  ylab = "Stage (m)"
)
```

The reported uncertainties give the variances directly, through `var_spec()`.
Below the breakpoint, $h < k$, the curve is a single power law,
$Q = a_1 (h - c_1)^{b_1}$. How the two segments combine above it, $h \ge k$,
is set by `combine`. With `"replace"` (the default), the upper power law
takes over from the lower one:

$$
Q = a_2 (h - c_2)^{b_2},
\qquad \text{where } a_2 = \frac{a_1 (k - c_1)^{b_1}}{(k - c_2)^{b_2}},
$$

so that the two meet at the breakpoint. With `"add"`, the upper power law
adds to the discharge the lower one carries at the breakpoint, as when flow
spills onto a floodplain:

$$
Q = a_1 (k - c_1)^{b_1} + a_2 (h - k)^{b_2}.
$$

```{r, eval = has_rbam}
fit_repl <- rc_2seg_powerlaw(
  discharge, stage,
  data = sauze,
  variance = var_spec(sauze$uncertainty_sd^2)
)
fit_add <- rc_2seg_powerlaw(
  discharge, stage,
  data = sauze,
  combine = "add",
  variance = var_spec(sauze$uncertainty_sd^2)
)
c(
  replace = fit_repl$curve_parameters$k,
  add = fit_add$curve_parameters$k
)
```

The fit is sensitive to where the search for the breakpoint starts, so by
default it tries several starting values and keeps the best. What each
start led to is recorded:

```{r, eval = has_rbam}
fit_repl$kstart_search
```

```{r, eval = has_rbam}
plot(stage ~ discharge, data = sauze, col = "grey50",
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
lines(stage ~ fit, data = predict(fit_repl), lwd = 2)
lines(stage ~ fit, data = predict(fit_add), lwd = 2, lty = 2)
abline(h = fit_repl$curve_parameters$k, col = "grey", lty = 3)
legend("bottomright", c("Upper segment replaces", "Upper segment adds"),
       lty = 1:2, lwd = 2, bty = "n")
```

Gaugings are sparse near the breakpoint at Sauze, so where exactly it lies is
uncertain, which affects the limits on the curve there; see
`vignette("uncertainty")`.
