---
title: "Uncertainty in predicted discharge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty in predicted discharge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4.2,
  out.width = "100%"
)
has_rbam <- requireNamespace("RBaM", quietly = TRUE)
```

A rating curve turns a recorded stage into a discharge, and that discharge is
uncertain. This vignette shows how CSHShydRometry puts limits on it, what the
limits mean, and where they can mislead. The fits themselves are covered in
`vignette("fitting")`, which also explains why the Thompson gaugings used here
illustrate the methods rather than give a rating for any one period.

```{r setup}
library(CSHShydRometry)
```

## Two kinds of limits

`predict()` gives two kinds of limits, each at a level you choose:

- **Confidence limits** (`conflev`) bound the *curve*: the mean discharge at
  a stage. They reflect only how well the gaugings pin the curve down, and
  narrow as gaugings accumulate.
- **Prediction limits** (`predlev`) bound a *single new gauging* at that
  stage. They add the scatter of individual gaugings about the curve, so they
  are always wider, and do not narrow below that scatter however many
  gaugings there are.

```{r}
fit <- rc_powerlaw(discharge, stage, data = thompson, variance = "prop")
predict(fit, new_stage = c(1, 4, 8), conflev = 0.95, predlev = 0.95)
```

Which to use depends on the question. Discharge averaged over a period, where
the scatter of individual gaugings largely cancels, calls for confidence
limits. Discharge at a particular moment calls for prediction limits.

```{r}
band <- predict(fit, conflev = 0.95, predlev = 0.95)

plot(stage ~ discharge, data = thompson, col = "grey50",
     xlab = "Discharge (m³/s)", ylab = "Stage (m)")
lines(stage ~ fit, data = band, lwd = 2)
lines(stage ~ ci_lwr, data = band, lty = 2)
lines(stage ~ ci_upr, data = band, lty = 2)
lines(stage ~ pi_lwr, data = band, lty = 3)
lines(stage ~ pi_upr, data = band, lty = 3)
legend("bottomright", c("Fitted curve", "95% confidence limits",
       "95% prediction limits"), lty = 1:3, lwd = c(2, 1, 1), bty = "n")
```

## The variance scheme shapes the limits

The variance scheme is a statement about the scatter, so it decides how the
prediction limits change with the flow. With `var_none()` the scatter is the
same at every flow; with `var_prop()` it grows in proportion to the flow.

```{r}
at <- c(1, 8)
widths <- function(fit) {
  p <- predict(fit, new_stage = at, predlev = 0.95)
  p$pi_upr - p$pi_lwr
}
data.frame(
  stage = at,
  none = widths(rc_powerlaw(discharge, stage, data = thompson)),
  prop = widths(fit)
)
```

Under `var_none()` the prediction band is nearly as wide at low flow as at
high flow, which is implausible here: the gaugings in `vignette("fitting")`
scatter in proportion to the flow. Under `var_prop()` the band is narrow at
low flow and wide at high flow, as the gaugings are.

Under `var_spec()`, each gauging's variance is supplied, up to a common
factor, rather than modelled, so the fit says nothing about the scatter of a
new gauging. Its prediction
limits come back as `NA`, while its confidence limits are computed as usual.

## What the limits assume

The limits use the t distribution, which assumes the gaugings scatter
normally about the curve. Confidence limits depend on this only mildly,
because the estimates average over many gaugings. Prediction limits depend
on it directly: they are only as good as the assumed shape of the scatter,
and if it is skewed or has heavy tails they can miss, especially at high
levels such as 0.99. So look at the residuals before relying on prediction
limits. The Pearson residuals, each divided by its modelled standard
deviation, are the ones to check. Here, for the power law under
`var_prop()`, they look roughly normal:

```{r, fig.height = 3.4}
pearson <- residuals(fit, type = "pearson")
qqnorm(pearson, main = "Pearson residuals")
qqline(pearson)
```

Even with normal scatter, a 95% interval from most fits covers the truth
only roughly 95% of the time, because the calculation approximates the
curve by a straight line about its estimates. The approximation is good with
plenty of gaugings, and poorer with few; see `?predict.rc_powerlaw`.

## Fits on the log scale

`rc_powerlaw_log()` works on the log scale, where the limits are symmetric.
Transformed back to discharge they are lopsided, stretching further above
the curve than below it:

```{r}
fit_log <- rc_powerlaw_log(discharge, stage, data = thompson)
p <- predict(fit_log, new_stage = 8, predlev = 0.95)
c(below = p$fit - p$pi_lwr, above = p$pi_upr - p$fit)
```

The curve itself estimates the geometric mean discharge, which is below the
arithmetic mean by a factor that grows with the scatter. With scatter of
about `r round(100 * fit_log$rse)`% here, the difference is small:

```{r}
fit_log$a_corrected / fit_log$curve_parameters$a
```

### The offset

When the offset, $c$, is estimated, its uncertainty is part of
the limits. Holding it fixed removes that part. That is right when $c$ is
known independently, say from a survey of the control. Fixing it at the value
estimated from the same gaugings, though, gives the same curve with limits
that are too narrow:

```{r}
fixed <- rc_powerlaw_log(discharge, stage, data = thompson,
                      offset = fit_log$curve_parameters$c)
width_at <- function(fit, at) {
  p <- predict(fit, new_stage = at, conflev = 0.95)
  p$ci_upr - p$ci_lwr
}
c(c_estimated = width_at(fit_log, 3), c_fixed = width_at(fixed, 3))
```

## Beyond the gaugings

Limits are only as good as the model, and the model is only checked where
there are gaugings. Beyond them, the choice of curve matters more than any
limit suggests. Here three curves that agree within the gaugings are taken
half again past the highest one:

```{r}
top <- max(thompson$stage)
beyond <- seq(top, 1.5 * top, length.out = 50)
fits <- list(
  "Power law" = fit,
  "Cubic polynomial" = rc_poly(discharge, stage, data = thompson, degree = 3),
  "Loess" = rc_loess(discharge, stage, data = thompson)
)
curves <- lapply(fits, predict, new_stage = beyond)

plot(NULL, xlim = range(sapply(curves, function(p) range(p$fit))),
     ylim = range(beyond), xlab = "Discharge (m³/s)", ylab = "Stage (m)")
for (i in seq_along(curves)) {
  lines(stage ~ fit, data = curves[[i]], lty = i, lwd = 2)
}
legend("bottomright", names(fits), lty = seq_along(fits), lwd = 2, bty = "n")
```

The power law follows the physics of flow through a control, so it is the
safer guide for modest extrapolation. A polynomial or loess curve has no such
basis outside the data. None of their limits accounts for the chance that
the control itself changes at higher stages.

## Two-segment curves: limits near the breakpoint

For a two-segment curve, `predict()` offers two methods. The default,
`method = "delta"`, approximates the curve by a straight line in its
parameters. That works well away from the breakpoint, but the curve has a
corner there, and the approximation jumps across it. `method = "boot"` refits
the curve to resampled gaugings instead, which makes no such approximation.

```{r, eval = has_rbam}
sauze <- data.frame(
  stage = RBaM::SauzeGaugings$H,
  discharge = RBaM::SauzeGaugings$Q,
  uncertainty_sd = RBaM::SauzeGaugings$uQ
)
fit2 <- rc_2seg_powerlaw(discharge, stage, data = sauze,
                      variance = var_spec(sauze$uncertainty_sd^2))
near_k <- fit2$curve_parameters$k + c(-0.05, 0.05)

delta <- predict(fit2, new_stage = near_k, conflev = 0.95)
boot <- predict(fit2, new_stage = near_k, conflev = 0.95,
                method = "boot", B = 200, seed = 1)
data.frame(
  stage = near_k,
  delta_width = delta$ci_upr - delta$ci_lwr,
  boot_width = boot$ci_upr - boot$ci_lwr
)
```

Just above the breakpoint the delta band is several times wider than just
below it, a jump that comes from the approximation, not the data. In a
simulation study, a nominal 95% delta interval covered the true curve only
about two-thirds of the time just above the breakpoint, against roughly 97%
for the bootstrap.

The bootstrap costs a full refit per resample, and the refits repeat the
search over starting breakpoints. So use `"delta"` for a quick look, and
`"boot"` where the limits near the breakpoint matter.
